The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-17 dictionary.tex
Except for Aut(L), Section 3 is complete.

\begin{proposition}\label{prop:mor_etale}
Let $G$ be an etale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\alpha\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

\begin{proposition}\label{prop:connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then
 \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
 Moreover, for every quasicharacter sheaf $\qcs{L}$ on $G$, $\Aut(\qcs{L}) = 1$.
\end{proposition}

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to bounded quasicharacter sheaf $\qcs{E}$ on $G$. Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by etale descent, pull-back by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every bounded quasicharacter sheaf $\qcs{E}$ on $G$. 
Note that since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_* V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constand sheaf $V$ on $G$, equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$.
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\begin{lemma}\label{lem:extension}
The group homomorphism $\pi_0^*: \QCiso{\pi_0(G)} \to \QCiso{G}$ is injective.
\end{lemma}
\begin{proof}
\end{proof}

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme over $\Fq$ then
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
is surjective with kernel isomorphic to $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$.
\end{theorem}

@@

%We now turn our attention $\lTrFrob{G}: \lQCiso{G}  \to G(\Fq)^*$ is an isomorphism.
%First, note that $\lTrFrob{G}$ is surjective because $\TrFrob{G}$ is surjective.
%We will now show that the kernel of $\lTrFrob{G}$ is trivial.

%First, suppose $G$ is etale.
%Let $(\gqcs{E},\phi')$ be a quasicharacter sheaf on $G$ with $\trFrob{(\gqcs{E},\phi')}=1$.
%Let $\qcs{E} = (\gqcs{E},\mu,\phi')$ be a quasicharacter sheaf.
%By Proposition~\ref{prop:etale-iso}, $S_G(\qcs{E}) = [\alpha\oplus 0]$ for some $\alpha \in Z^0(\Weil{},Z^2({\bar G},\EEx))$
%corresponding to a choice $\{ v_x \in \gqcs{E}_x^\times \tq x\in {\bar C}\}$.
%Let $a\in Z^2({\bar G},\EEx)$ be the multiplicative version of $\alpha$
%and let $\mu'$ be the indexed set of isomorphisms $\mu'_{x,y} : \EE \to \EE\otimes \EE$ defined by $1 \mapsto a(x,y) 1 \otimes 1$.
%Then $(\EE,\mu', 1)$ is a quasicharacter sheaf the isomorphisms $\gqcs{E}_x \to \EE$ defined by $v_x \mapsto 1$ gives an isomorphism $(\gqcs{E},\mu,\phi')\to ((\EE)_{\bar C},\mu', 1)$ in $\QC(G)$.
%It follows that $(\gqcs{E},\phi')\iso ((\EE)_{\bar G},1)$ in $\lQC(G)$, so $\lTrFrob{G}$ is injective.
%This completes the proof of Theorem~\ref{thm:snake} in the case that $G$ is etale.

%Now let $G$ be arbitrary
%and let $(\gqcs{L},\phi)$ be a flabby quasicharacter sheaf on $G$ with $\trFrob{(\gqcs{L},\phi)}=1$.
%Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf.
%We saw above that $\ker \TrFrob{\pi_0(G)}\to \TrFrob{G}$ is surjective in this case,
%so $\qcs{L} = \pi_0^*\qcs{E}$ for some $\qcs{E}\in \QC(\pi_0(G))$.
%Since $\trFrob{\qcs{L}}(g) = \trFrob{\qcs{E}}(\pi_0(g))$ and since $\pi_0 : G(\Fq) \to \pi_0(G)(\Fq)$ is surjective, $\trFrob{\qcs{E}}=1$.
%But now $\qcs{E} \iso ((\EE)_{\pi_0({\bar G})},\mu', 1)$ by the paragraph above.
%So $\qcs{L} = \pi_0^*\qcs{E} \iso \pi_0^*((\EE)_{\pi_0({\bar G})},\mu', 1) = ((\EE)_{\bar G},\pi_0^*\mu', \pi_0^*1)$ in $\QC(G)$.
%It follows that $(\gqcs{L},\phi) \iso ((\EE)_{\bar G},1)$ in $\lQC(G)$.
%This completes the proof of Theorem~\ref{thm:snake}.
\end{proof}

%\begin{corollary}\label{cor:exact}
%  If $G$ is a smooth, commutative group scheme over
%  $\Fq$ then pullback yields an exact sequence
%\[
%0 \to \lQCiso{\pi_0(G)} \to \lQCiso{G} \to \lQCiso{G^0} \to 0
%\]
%of abelian groups.
%\end{corollary}

%\begin{proof}
%Argue as in the first paragraph in the proof of Theorem~\ref{thm:snake},
%using the category $\lQC(G)$ in place of $\QC(G)$.
%\end{proof}

\subsection{Morphisms, in the general case}

>>>>

\begin{proposition}\label{prop:mor_etale}
Let $G$ be an etale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

\begin{proposition}\label{prop:connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then
 \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
 Moreover, for every quasicharacter sheaf $\qcs{L}$ on $G$, $\Aut(\qcs{L}) = 1$
 and every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
 \end{proposition}

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to bounded quasicharacter sheaf $\qcs{E}$ on $G$. Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by etale descent, pull-back by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every bounded quasicharacter sheaf $\qcs{E}$ on $G$. 
Note that since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_* V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constand sheaf $V$ on $G$, equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$.
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
For the last part of the proposition, argue as in the analogous part of Proposition~\ref{prop:mor_etale}.
\end{proof}

@@

\begin{lemma}\label{lem:extension}
The group homomorphism $\pi_0^*: \QCiso{\pi_0(G)} \to \QCiso{G}$ is injective.
\end{lemma}
\begin{proof}
Let $\qcs{L}$ be a quasicharacter sheaf on $\pi_0(G)$ and let $\rho : \pi_0^*\qcs{L} \to (\EE)_{G}$ be an isomorphism in $\QC(G)$. For each $x\in \pi_0(\bG)$, the restriction $\pi_0^*\gqcs{L}\vert_{\bG^x}$ is the constant sheaf $(\gqcs{L}_x)_{\bG^x}$ so the isomorphism $\rho_{\bG^x} : (\gqcs{L}_x)_{\bG^x} \to (\EE)_{\bG^x}$ determines an isomorphism $\rho_x : \gqcs{L}_x \to (\EE)_x$. The collection $\{ \rho_x \tq x\in \pi_0(\bG) \}$ determines an isomorphism $\qcs{L} \to (\EE)_{\pi_0(G)}$  in $\QC(\pi_0(G))$.
\end{proof}

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme over $\Fq$ then
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
is surjective with kernel isomorphic to $\Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$.
 Moreover, $\Aut(\qcs{L}) = \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism, for all quasicharacter sheaves $\qcs{L}$ and $\qcs{L}'$ on $G$. 
\end{theorem}

@@

 (. . . where $G(\bFq)$ is given the topology of the espace étalé of the sheaf of sections of $G$.)
 \todo{Still under construction: $\Aut(\qcs{L})$.}

 For the last part of the proposition, argue as in the analogous part of Proposition~\ref{prop:mor_etale}.
\end{proof}
